\chapter{\Conc\ model: implementation}\label{ch7}

\begin{zsection}
  \SECTION ch7 \parents ch6
\end{zsection}

\section{Concrete World State}

The \Conc\ world state has the same components as the \Betw\ state;
we decorate with a subscript zero to distinguish like-named \Betw\
and \Conc\ components.

Since $\Delta ConWorld_0$ has components dashed-then-subscripted,
whereas we require subscripted-then-dashed, Mondex defines their own
$\Delta$ and $\Xi$ schemas. as
\newcommand{\DeltaConWorldO}{\Delta ConWorld0}
\newcommand{\XiConWorldO}{\Xi ConWorld0}
%
\begin{gzed}
  \DeltaConWorldO ~~\defs~~ ConWorld_0 \land ConWorld_0~' \\
  \XiConWorldO ~~\defs~~ [~ \DeltaConWorldO | \theta ConWorld_0 = \theta ConWorld_0~' ~]
\end{gzed}
%
Nevertheless, there are two curious subtleties regarding this
definition:~(i) an oddity in the Z standard;~and~(ii)~tool
limitation. The curiosity in the Z standard is that in \LaTeX\
\verb+x_0'+ ($x_0'$) typesets exactly the same as \verb+x'_0+. For
this there is an easy solution: \verb+x_0~'+ ($x_0~'$). On the other
hand, for \zeves, when decorated names are used in the way we need
($x_0~'$), the $\pre$ expression does not work as expected, as it
does not recognised that all subscribed-primed components need to be
existentially quantified. This introduces a complication in the
precondition calculation in Chapter~\ref{ch8} for all \Conc\ model
operations. Therefore, we decided to use variable renaming instead
of subscripting here. It would be possible to keep the original, but
this would incur in error prone preconditions done by hand for
Chapter~\ref{ch8} and all refinement proofs. In the end, there is no
semantic difference as this is just a ``syntactic cosmetic
surgery''.

\begin{LNewSDef}[Renamed $ConWorld$ for Chapter~7]
\begin{zed}
    ConWorldCh7 \defs ConWorld[conAuthPurseCh7/conAuthPurse, \\
                \t1 etherCh7/ether, archiveCh7/archive]
\end{zed}~\end{LNewSDef}

\section{Framing Schema}\label{ch7.PhiCOp}

\begin{LSDef}[Concrete world promotion]
\begin{schema}{PhiCOp}
  \Delta ConWorldCh7
  \\      \Delta ConPurse
  \\      m?,m! : MESSAGE
  \\      name? : NAMES
  \where
  m? \in etherCh7
  \also
  name? \in \dom~conAuthPurseCh7
  \\      \theta ConPurse = conAuthPurseCh7~name?
  \also
  conAuthPurseCh7' = conAuthPurseCh7 \oplus \{~ name? \mapsto \theta ConPurse~' ~\}
  \also
  archiveCh7' = archiveCh7
  \also
  etherCh7' \subseteq etherCh7 \cup \{~m!~\}
\end{schema}~\end{LSDef}

\section{Ignore, Increase and Abort}

In the original the next declarations were given in wrong \LaTeX\ format. It
did not followed Spivey's grammar which enforces one declaration per unboxed
paragraph. That means we need a \verb|begin{zed}| \verb|end{zed}| environment
for each of definition, rather than one for all definitions.
%
\begin{LSDef}[Concrete world promoted ignore operation]
\begin{zed}
  CIgnore~~\defs~~[~ \Xi ConWorldCh7; name? : NAMES; m?,m! : MESSAGE | m! = \bot ~]
\end{zed}~\end{LSDef}

\begin{LSDef}[Concrete world promoted increase operation]
\begin{zed}
  CIncrease~~\defs~~CIgnore
  \\ \t2  \lor (~ \exists \Delta ConPurse @ PhiCOp \land IncreasePurseOkay ~)
\end{zed}~\end{LSDef}
%
\begin{LSDef}[Concrete world promoted abort operation]
\begin{zed}
  CAbort~~\defs~~CIgnore
  \\ \t2          \lor (~\exists \Delta ConPurse @
  AbortPurseOkay \land [~ PhiCOp | m! = \bot ~]~)
\end{zed}~\end{LSDef}

\section{Promoted operations}

\subsection{Value transfer operations}

\begin{LSDef}[$ConWorld$ promoted $startFrom$ operation]
\begin{zed}
  CStartFrom~~\defs~~CIgnore
  \\ \t2          \lor CAbort
  \\ \t2          \lor (~ \exists \Delta ConPurse @
  PhiCOp \land StartFromPurseOkay ~)
\end{zed}~\end{LSDef}

\begin{LSDef}[$ConWorld$ promoted $startTo$ operation]
\begin{zed}
  CStartTo~~\defs~~CIgnore
  \\ \t2          \lor CAbort
  \\ \t2          \lor (~ \exists \Delta ConPurse @
  PhiCOp \land StartToPurseOkay ~)
\end{zed}~\end{LSDef}

Same \LaTeX\ unboxed environment problem sorted here.
%
\begin{LSDef}[$ConWorld$ promoted request operation]
\begin{zed}
  CReq~~\defs~~CIgnore \lor (~ \exists \Delta ConPurse @
  PhiCOp \land ReqPurseOkay ~)
\end{zed}~\end{LSDef}

\begin{LSDef}[$ConWorld$ promoted value transfer operation]
\begin{zed}
  CVal~~\defs~~CIgnore \lor (~ \exists \Delta ConPurse  @
  PhiCOp \land ValPurseOkay ~)
\end{zed}~\end{LSDef}

\begin{LSDef}[$ConWorld$ promoted acknowledgement operation]
\begin{zed}
  CAck~~\defs~~CIgnore \lor (~ \exists \Delta ConPurse @
  PhiCOp \land AckPurseOkay ~)
\end{zed}~\end{LSDef}

\subsection{Exception log operations}

\begin{LSDef}[$ConWorld$ promoted read exception log operation]
\begin{zed}
  CReadExceptionLog~~\defs~~CIgnore
  \\ \t2           \lor (~ \exists \Delta ConPurse @
  PhiCOp \land ReadExceptionLogPurseOkay ~)
\end{zed}~\end{LSDef}

\begin{LSDef}[$ConWorld$ promoted clear exception log operation]
\begin{zed}
  CClearExceptionLog~~\defs~~CIgnore
  \\ \t2          \lor CAbort
  \\ \t2          \lor (~ \exists \Delta ConPurse @
  PhiCOp \land ClearExceptionLogPurseOkay ~)
\end{zed}~\end{LSDef}

\section{Operations at the world level only}

\subsection{Exception Log clear authorisation}

\begin{LNewSDef}[$ConWorld$ promoted exception log clear]
\begin{zed}
   CPhiCOpExpLogClearNEW ~~\defs~~ [~ PhiCOp |
  (~\exists pds : \power_1 PayDetails |
  \\ \t4                          \{~ name? ~\} \cross pds \subseteq archiveCh7 @
  \\ \t5                                  m! = exceptionLogClear~(name?, image~pds ) ~)
  ~]
\end{zed}~\end{LNewSDef}

\begin{LSDef}[$ConWorld$ promoted authorise exception log clear]
\begin{zed}
  CAuthoriseExLogClear~~\defs~~CIgnore \lor (~ \exists \Xi ConPurse @ CPhiCOpExpLogClearNEW~)
\end{zed}~\end{LSDef}

The same \LaTeX\ problems for subscripted names occurs here.
%
\begin{LSDef}[Exception log archiving operation]
\begin{schema}{CArchive}
  \Delta ConWorldCh7
  \\      m?, m!: MESSAGE
  \\      name? : NAMES
  \where
  conAuthPurseCh7' = conAuthPurseCh7
  \\      etherCh7' \subseteq etherCh7
  \also
  archiveCh7 \subseteq
\\ \t1          archiveCh7' \subseteq
  \\ \t2          archiveCh7' \cup \{~ log: NAMES \cross PayDetails |
  \\ \t5                          exceptionLogResult~log \in etherCh7 ~\}
  \also
  m! = \bot
\end{schema}~\end{LSDef}

\section{Initial state}

The same \LaTeX\ problems for subscripted names occurs here.
%
\begin{LSDef}[$ConWorld$ initialisation state]
\begin{schema}{ConInitState}
  ConWorldCh7~'
  \where
  \exists BetweenWorld~' | BetweenInitState @
  \\ \t1          conAuthPurseCh7' = conAuthPurse'
  \\ \t1          \land archiveCh7' = archive'
  \\ \t1          \land \{~\bot~\} \subseteq etherCh7' \subseteq ether'
\end{schema}~\end{LSDef}

\section{Finalisation}

In here the same finiteness problem that occurs in $BetwFinState$
(Section~\ref{ch6.final} on page \pageref{ch6.final}) (Section $6.2$ on page
$53$ on original document).

\begin{LNewSDef}[Renamed $AuxWorld$ for Chapter~7]
\begin{zed}
    AuxWorldCh7 ~~\defs~~ AuxWorld[conAuthPurseCh7/conAuthPurse, \\
        \t1 etherCh7/ether, archiveCh7/archive, \\
        \t1 allLogsCh7/allLogs, authenticFromCh7/authenticFrom, \\
        \t1 authenticToCh7/authenticTo, fromLoggedCh7/fromLogged, \\
        \t1 toLoggedCh7/toLogged, toInEpvCh7/toInEpv, \\
        \t1 toInEapayeeCh7/toInEapayee, fromInEprCh7/fromInEpr, \\
        \t1 fromInEpaCh7/fromInEpa, definitelyLostCh7/definitelyLost, \\
        \t1 maybeLostCh7/maybeLost]
\end{zed}~\end{LNewSDef}

\begin{LSDef}[$ConWorld$ finalisation state]
\begin{schema}{ConFinState}
  AuxWorldCh7
  \\ %
  GlobalWorld
\where %
  \dom gAuthPurse = \dom conAuthPurseCh7
  \also %
%  \znote{Obvious assumption that the set comprehension}
%  \znote{below in the set intersection is finite.}
  \forall name: \dom~conAuthPurseCh7 @
  \t1 (gAuthPurse~name).balance
  \\ %
  \t1 = (conAuthPurseCh7~name).balance
  \also %
  \t1 \land (gAuthPurse~name).lost =
  \\ %
  \t2 sumValue~( (definitelyLostCh7 \cup maybeLostCh7)
  \\ %
  \t3 \cap \{~ ld:PayDetails | ld.from = name ~\})
\end{schema}~\end{LSDef}


\newpage

\section{Summary}\label{ch7.summary}

\ldefsummary %
\lthmsummary %
\lthmaddeddefsummary %
\lthmaddedthmsummary %
\lzevessummary %
%\lpscriptsummary
