\documentclass[11pt]{article}
\usepackage{ltcadiz}

\begin{document}

\begin{zsection}
  \SECTION mtr \parents oz\_toolkit, standard\_toolkit
\end{zsection}

\zedsize{\small}
\begin{class}{Clock}
\also
\begin{state}
date:  \nat
\end{state}\classbreak
\begin{init}
date = 0
\end{init}\classbreak
\begin{op}{getDate}
today!:  \nat  
\where
today! = date
\end{op}\classbreak
\begin{op}{nextDay}
\Delta (date)
\where
date' = date + 1
\end{op} 
\end{class} 

\begin{class}{FareDataBase[STATION]}
\also
\begin{state}
fare: STATION \cross STATION \pfun  \nat  
\end{state} \classbreak
\begin{init}
fare = \emptyset
\end{init} \classbreak
\begin{op}{addFare}
\Delta (fare)\\
start?: STATION \\ 
finish?: STATION \\ 
fare?:  \nat  
\where
(start?, finish?) \notin  \dom fare \\ 
fare' = fare \cup \{(start?, finish?) \mapsto fare?\} \\ 
\end{op} \classbreak
\begin{op}{updateFare}
\Delta (fare)\\
start?: STATION \\ 
finish?: STATION \\ 
newfare?:  \nat  
\where
 (start?, finish?) \in  \dom fare \\ 
fare' = fare \oplus \{(start?, finish?) \mapsto newfare?\} \\ 
\end{op} \classbreak 
\begin{op}{checkFare}
start?: STATION \\ 
finish?: STATION \\ 
fare!:  \nat  
\where
 (start?, finish?) \in  \dom fare \\ 
fare! = fare(start?, finish?)
\end{op} 
\end{class} 


\begin{class}{Tickets[TICKET, STATION]}
\also
%TICKET == 1 \upto 10\\
%STATION == 1 \upto 10\\
TYPE ::= single | multi | season \\
MultiValue == \{50, 100\} \\ 
SeasonDuration == \{7, 31, 365\}\\
\begin{state}
singles:  \power TICKET \\ 
multis:  \power TICKET \\ 
seasons:  \power TICKET \\ 
value: TICKET \pfun  \nat  \\ 
expiry: TICKET \pfun  \nat  \\ 
entrypoint: TICKET \pfun STATION 
\where
 \dom value = (singles \cup multis \cup seasons) \\ 
\forall t : (singles \cup multis \cup seasons) @
   \# (\{t\} \dres entrypoint) \leq 1
\end{state} \classbreak 
\begin{init}
expiry = \emptyset \\ 
singles = \emptyset \\ 
multis = \emptyset \\ 
seasons = \emptyset \\ 
value = \emptyset \\ 
entrypoint = \emptyset
\end{init} \classbreak 
\begin{op}{issue}
\Delta (expiry, singles, multis, seasons, value)\\
ticket?: TICKET \\ 
today?:  \nat  \\ 
type?: TYPE \\ 
value?:  \nat  
\where
 ticket? \notin (singles \cup multis \cup seasons) \land\\
       (type? = multi \implies value? \in MultiValue) \land\\
       (type? = season \implies value? \in SeasonDuration)\\
type? = single \implies \\
\t1 expiry' = expiry \cup \{ticket? \mapsto today?\} \land \\ 
\t1 singles' = singles \cup \{ticket?\} \land \\ 
\t1 value' = value \cup \{ticket? \mapsto value?\} \land\\ 
\t1 multis' = multis \land seasons' = seasons\\
type? = multi \implies\\
\t1 expiry' = expiry \cup \{ticket? \mapsto (today? + 730)\} \land \\ 
\t1 value' = value \cup \{ticket? \mapsto value?\} \land \\ 
\t1 multis' = multis \cup \{ticket?\} \land \\ 
\t1 singles' = singles \land seasons' = seasons\\
type? = season \implies\\
\t1 expiry' = expiry \cup \{ticket? \mapsto today? + value?\} \land \\ 
\t1 value' = value \cup \{ticket? \mapsto 1\} \land \\ 
\t1 seasons' = seasons \cup \{ticket?\} \land \\ 
\t1 singles' = singles \land  multis' = multis
\end{op}
\classbreak 
\begin{op}{reIssue}
\Delta (expiry, value, entrypoint)\\
ticket?: TICKET \\ 
today?:  \nat  \\ 
value?:  \nat  
\where
 ticket? \in (singles \cup multis \cup seasons) \land\\
     (ticket? \in multis \implies value? \in MultiValue) \land\\
     (ticket? \in seasons \implies value? \in SeasonDuration)\\
ticket? \in singles \implies\\
\t1 expiry' = expiry \oplus \{ticket? \mapsto today?\} \land \\ 
\t1 value' = value \oplus \{ticket? \mapsto value?\}\\ 
ticket? \in multis \implies\\
\t1 expiry' = expiry \oplus  \{ticket? \mapsto (today? + 730)\} \land \\ 
\t1 value' = value \oplus \{ticket? \mapsto value?\} \\ 
ticket? \in seasons \implies\\
\t1 expiry' = expiry \oplus \{ticket? \mapsto today? + value?\} \land \\ 
\t1 value' = value \oplus \{ticket? \mapsto 1\}\\
entrypoint' = \{ticket?\} \ndres entrypoint
\end{op} \classbreak 
\begin{op}{enterStation}
\Delta (entrypoint)\\
ticket?: TICKET \\ 
station?: STATION \\ 
today?:  \nat  
\where
 ticket? \in (singles \cup multis \cup seasons) \land \\ 
\t2 value(ticket?) \neq 0 \land \\ 
\t2 expiry(ticket?) \geq today? \land \\ 
\t2 \{ticket?\} \dres entrypoint = \emptyset \\ 
entrypoint' = entrypoint \cup \{ticket? \mapsto station?\} \\ 
\end{op} \classbreak 
\begin{op}{entry}
ticket?: TICKET \\ 
station!: STATION 
\where
 ticket? \in  \dom entrypoint \\ 
station! = entrypoint(ticket?) \\ 
\end{op} \classbreak 
\begin{op}{exitStation}
\Delta (entrypoint, value)\\
fare?:  \nat  \\ 
ticket?: TICKET \\ 
start!: STATION 
\where
 ticket? \in (singles \cup multis \cup seasons) \land \\ 
ticket? \in  \dom entrypoint \\ 
ticket? \in singles \implies\\
\t1 fare? \leq value(ticket?) \land \\ 
\t1 value' = value \oplus \{ticket? \mapsto 0\}\\
ticket? \in multis \implies\\ 
\t1 value' = value \oplus \{ticket? \mapsto max \{0, value(ticket?) - fare?\}\}\\ 
ticket? \in seasons \implies\\
\t1 value' = value\\
start! = entrypoint(ticket?) \\ 
entrypoint' = \{ticket?\} \ndres entrypoint
\end{op} 
\end{class} 
\end{document}
